Analyzing the effects of COVID-19 related social media use on well-being

### 2022-04-12 manuscript.tex
- implented reviewer feedback

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,mask,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Analyzing the effects of COVID-19 related social media use on well-being},
  pdflang={en-EN},
  pdfkeywords={COVID-19, Coronavirus, well-being, affect, life satisfaction, social media use, news use, communication, random effects within between model, panel study, longitudinal study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

@@

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\keywords{COVID-19, Coronavirus, well-being, affect, life satisfaction, social media use, news use, communication, random effects within between model, panel study, longitudinal study.}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\setlength{\parskip}{0em}
\raggedbottom
\note{\clearpage}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[main=english]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Analyzing the effects of COVID-19 related social media use on well-being}
\author{Name blinded for review\textsuperscript{1}}
\date{}

\shorttitle{COVID-19 related social media use and well-being}

\authornote{

Correspondence concerning this article should be addressed to Name blinded for review, . E-mail:

}

@@

\abstract{
In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, the latest political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop `doomscrolling.' In this preregistered study, I investigate whether using social media for COVID-19 related reasons affects psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. The findings suggest that fears that social media use during times of crisis impairs well-being are likely to be unfounded.
}

@@

Some people actually couldn't stop using social media to learn about COVID-19 related news.
A new phenomenon emerged, termed ``doomscrolling'':
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children (Klein, 2021).
As doomscrolling increased it became doubtful whether such social media is helpful, or whether it created an additional burden on mental health (Sandstrom, Buchanan, Aknin, \& Lotun, 2021).
These concerns seem justified:
A study with 6,233 people from Germany conducted during the pandemic found that ``{[}f{]}requency, duration and diversity of media exposure were positively associated with more symptoms of depression'' (Bendau et al., 2021, p. 283).

@@

First, I investigate how different types of communication affect well-being.
Specifically, I differentiate between active and passive use.
I distinguish (a) \emph{reading} (passive), (b) \emph{posting} (active), and (c) \emph{liking and sharing} COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects.
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, Waterloo, Baumgartner, Peter, and Valkenburg (2018) found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about \emph{general} social media use during times of COVID, but on social media use \emph{focused} on COVID-19 related content.
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

\hypertarget{effects-of-social-media-on-well-being}{%
\subsection{Effects of Social Media on Well-Being}\label{effects-of-social-media-on-well-being}}

So far, there is only little empirical research on COVID-19 related social media use on well-being.
In their study on the relations between media use and mental health during the pandemic, Bendau et al. (2021) found that people who used social media as a primary source of information reported on average ``significantly more unspecific anxiety and depression {[}{]} and significantly more specific COVID-19 related anxiety symptoms'' (p.~288).
Eden, Johnson, Reinecke, and Grady (2020) analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, Choi and Choung (2021) reported that people who used media to attain information were more lonely and less satisfied with their lives.
Stainback, Hearne, and Trieu (2020) analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations (Sewall, Goldstein, \& Rosen, 2021).
The results showed that digital technology did not have any significant effects on mental health (for a similar study with comparable results, see Bradley \& Howard, 2021).
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use (see also Dörnemann, Boenisch, Schommer, Winkelhorst, \& Wingen, 2021; Liu \& Tong, 2020; Riehm et al., 2020).
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

@@

These overviews are well aligned with several individual studies employing advanced methods (Keresteš \& Štulhofer, 2020; Orben, Dienlin, \& Przybylski, 2019; Przybylski, Nguyen, Law, \& Weinstein, 2021; Schemer, Masur, Geiß, Müller, \& Schäfer, 2021).
For example, Beyens, Pouwels, Driel, Keijsers, and Valkenburg (2021) reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
In conclusion, most effects are likely somewhere between trivial and small.
I therefore expect that also in the case of COVID-19 related social media use effects will be trivial to small.

From a theoretical perspective, how could we explain whether COVID-19 related social media use might affect well-being?
In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent.
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.

First, one could assume a \emph{direct} negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing (Dörnemann, Boenisch, Schommer, Winkelhorst, \& Wingen, 2021), perhaps even changing some general perspectives on life.
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects (Dörnemann, Boenisch, Schommer, Winkelhorst, \& Wingen, 2021).

There could also be \emph{indirect} effects.
When doomscrolling, users are captivated to such an extent that they cannot stop using social media.
For example, during the pandemic social media use was at an all-time high in the US (Statista, 2021).
In general, as has been expressed by many before, it is most likely that moderate social media use is not detrimental (Orben, 2020).
Overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users (Przybylski \& Weinstein, 2017).
To explain, overuse likely impairs well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
So if a society collectively overuses social media when doomscrolling, there is potential for negative effects.

On the other hand, one can make the case that overuse might also be beneficial, especially in times of a pandemic---even if the use is mainly COVID-19 related.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.

Together, the strongest argument to me is that \emph{in general} the effects of social media on well-being are, on average, small at best.
Because this study only looks at \emph{one part} of social media use---namely, COVID-19 related interactions---it is very focused, diminishing the overall potential of the effects even further.
Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear.
Therefore, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.

@@

The sample was representative of the Austrian population in terms of age, gender, region/state, municipality size, and educational level.
In order to participate in the study, the respondents needed to be Austrian residents and had to be at least 14 years of age.
The study received IRB approval from the University of Vienna.
The average age was 42 years, 49 percent were male, 14 percent had a University degree, and 5 percent were currently unemployed.

@@

In this study, life satisfaction was measured on an 11-point scale.
If people can reliably differentiate 7 levels as mentioned above, this corresponds to 11 / 7 = 1.57 unit change on an 11-point scale.
Hence, a four-point change in media use (e.g., a complete stop) should result in a 1.57-point change in life satisfaction.
In a statistical regression analysis, \emph{b} estimates the change in the dependent variable if the independent variable increases by one point.
We would therefore expect a SESOI of \emph{b} = 1.57 / 4 = 0.39.
For affect, which was measured on a 5-point scale, our SESOI would be \emph{b} = 0.71 / 4 = 0.18.
Because we're agnostic as to whether the effects are positive or negative, the null region includes negative and positive effects.
Finally, in order not to exaggerate precision and to be less conservative, these numbers are reduced to nearby thresholds.\footnote{Note that other researchers also decreased or recommended decreasing thresholds for effect sizes when analyzing within-person or cumulative effects (Beyens, Pouwels, Driel, Keijsers, \& Valkenburg, 2021; Funder \& Ozer, 2019).}
Together, this leads to a null region ranging from \emph{b} = -.30 to \emph{b} = .30 for life satisfaction, and \emph{b} = -.15 to \emph{b} = .15 for positive and negative affect.

Let's briefly illustrate what this means in practice.
If the 95\% confidence interval falls completely within the null-region (e.g., \emph{b} = .20, {[}95\% CI: .15, .25{]}), the hypothesis that the effect is trivial is supported.
If the confidence interval and the null region overlap (e.g., \emph{b} = .30, {[}95\% CI: .25, .35{]}), the hypothesis is not supported and the results are considered inconclusive, while a meaningful negative effect is rejected.
If the confidence interval falls completely outside of the null-region (e.g., \emph{b} = .40, {[}95\% CI: .35, .45{]}), the hypothesis is rejected and the existence of a meaningful positive effect is supported.

@@

In what follows, I list all the variables that I analyzed.
For the variables' means, range, and variance, see Table \ref{tab:descriptives}.
For a complete list of all items and item characteristics, see \href{https://XMtRA.github.io/Austrian_Corona_Panel_Project}{companion website}.

@@

Life satisfaction was measured with the item ``Taken everything together, how satisfied are you currently with your life?''
The response options ranged from 0 (\emph{extremely unsatisfied}) to 10 (\emph{extremely satisfied}).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,mask]{apa6}
\title{Analyzing the effects of COVID-19 related social media use on well-being}
\author{Name blinded for review\textsuperscript{1}}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Analyzing the effects of COVID-19 related social media use on well-being},
  pdfauthor={Name blinded for review1},
  pdflang={en-EN},
  pdfkeywords={COVID-19, well-being, social media, news use, panel study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{COVID-19 related social media use and well-being}
\keywords{COVID-19, well-being, social media, news use, panel study.}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{endnotes}
\setlength{\parskip}{0em}
\raggedbottom
\note{\clearpage}
\ifXeTeX
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[main=english]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\abstract{
In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect. In this preregistered study, I investigate whether using social media for COVID-19 related reasons affects psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 3,018 participants. Well-being was measured at each wave, and communication at five waves. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. Fears that social media use during times of crisis critically impairs well-being are likely to be unfounded.
}

@@

Some people actually couldn't stop using social media to learn about COVID-19 related news.
A new phenomenon emerged, termed ``doomscrolling'':
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children (Klein, 2021).
It was asked whether using social media for COVID-19 related reasons is helpful, or whether it creates an additional burden on mental health (Sandstrom, Buchanan, Aknin, \& Lotun, 2021).
These concerns seem justified:
A study with 6,233 people from Germany conducted during the pandemic found that ``{[}f{]}requency, duration and diversity of media exposure were positively associated with more symptoms of depression'' (Bendau et al., 2021, p. 283).

@@

First, I investigate how well-being is affected by different types of communication affect, namely active and passive use.
Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed Meier \& Reinecke (2020).
Reading is generally considered as passive and writing as active, while there are also specific behaviors such as such liking or sharing content that fall somewhere in-between (Meier \& Krause, 2022).
In this study, I hence distinguish (a) \emph{reading} (passive), (b) \emph{posting} (active), and (c) \emph{liking and sharing} COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects.
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, Waterloo, Baumgartner, Peter, and Valkenburg (2018) found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about \emph{general} social media use during times of COVID, but on social media use \emph{focused} on COVID-19 related content.
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

\hypertarget{theorizing-social-media-effects-on-well-being}{%
\subsection{Theorizing Social Media Effects on Well-Being}\label{theorizing-social-media-effects-on-well-being}}

From a theoretical perspective, how could we explain whether COVID-19 related social media use might affect well-being?
According to the set-point model of subjective well-being, well-being is surprisingly stable (Lykken, 1999).
Although specific events such as marriage or salary can have significant effects, after some time well-being routinely returns to prior levels, which are mostly determined genetically (Sheldon \& Lucas, 2014).
Only very specific events such as unemployment, disability, or death can cause long-term decreases in well-being (Lucas, 2007).
So although well-being can change this does not happen easily.

Can media use be such a negative or positive factor?
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.
Empirically, social media use---on average---does not have a strong effect on well-being (Meier \& Reinecke, 2020).
According to the different susceptibility of media effects model (Valkenburg \& Peter, 2013), the effects of media use differ across individuals.
Whereas for some media are beneficial, for others they are harmful.
On average, however, effects are often small or negligible.
Also when focusing on individual users, social media have both positive and negative effects on well-being (Büchi, 2021).
They can impair well-being when causing embarrassment, stress, or disinformation, and they can improve well-being when providing connectedness, information, or entertainment (Büchi, 2021).

Two of the most prominent media effect theories argue against strong negative impacts.
First, uses and gratifications theory states that people explicitly and rationally chose specific media because of their respective benefits.
If those benefits don't exist, they invest their time elsewhere.
And social media offer ample benefits.
Most prominently, they help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself (Pelletier, Krallman, Adams, \& Hancock, 2020).
The second major theory is mood management theory (Zillmann, 1988).
Users implicitly learn what media help them balance their mood and affect.
For example, when bored people use social media to entertain themselves.
Using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased for a short time (Marciano, Driver, Schulz, \& Camerini, 2022).

Precisely because social media have so many positive consequences, one can ask if this is the main problem.
In other words, social media aren't problematic because they're inherently bad, but rather because they're too good.
And as with many other things, there can be too much of a good thing.
It is therefore often asked whether social media are addictive, and users sometimes express this fear themselves (Yang, Griffiths, Yan, \& Xu, 2021).
However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being (Duradoni, Innocenti, \& Guazzini, 2020).
In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social and new media represent new regular behaviors that should not be pathologized (Galer, 2018; van Rooij et al., 2018).

Because effects can differ across situations, I now briefly focus on the effects of COVID-19 related social media use specifically.
First, one could assume a \emph{direct} negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing (Dörnemann, Boenisch, Schommer, Winkelhorst, \& Wingen, 2021), perhaps even changing some general perspectives on life.
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects (Dörnemann et al., 2021).
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased during the pandemic (Guazzini, Pesce, Marotta, \& Duradoni, 2022).

There could also be \emph{indirect} effects.
When browsing social media for Covid-19 related news, many users reported being captivated to such an extent that they could not stop using social media (Klein, 2021).
During the pandemic social media use was at an all-time high in the US (Statista, 2021).
It is most likely that moderate social media use is not detrimental (Orben, 2020).
Overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users (Przybylski \& Weinstein, 2017).
To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
So if a society collectively overuses social media during a pandemic, there might be potential for negative effects.
On the other hand, one can make the case that overuse might also be beneficial, especially in times of a pandemic---even if the use is mainly COVID-19 related.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.

To summarize, from a theoretical perspective it is most likely that the average effects of social media use on well-being are negligible.
Building on established theories from Communication, it is not particularly likely that effects are either profoundly negative or strongly positive.

\hypertarget{empirical-studies-on-social-media-effects}{%
\subsection{Empirical Studies on Social Media Effects}\label{empirical-studies-on-social-media-effects}}

So far, there is only little empirical research on how well-being is affected by COVID-19 related social media.
In their study on the relations between media use and mental health during the pandemic, Bendau et al. (2021) found that people who used social media as a primary source of information reported on average ``significantly more unspecific anxiety and depression {[}{]} and significantly more specific COVID-19 related anxiety symptoms'' (p.~288).
Eden, Johnson, Reinecke, and Grady (2020) analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, Choi and Choung (2021) reported that people who used media to attain information were more lonely and less satisfied with their lives.
Stainback, Hearne, and Trieu (2020) analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations (Sewall, Goldstein, \& Rosen, 2021).
The results showed that digital technology did not have any significant effects on mental health (for a similar study with comparable results, see Bradley \& Howard, 2021).
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use (see also Dörnemann et al., 2021; Liu \& Tong, 2020; Riehm et al., 2020).
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

@@

Several individual studies employing advanced methods found smalls relations between social media use and well-being (Keresteš \& Štulhofer, 2020; Orben, Dienlin, \& Przybylski, 2019; Przybylski, Nguyen, Law, \& Weinstein, 2021; Schemer, Masur, Geiß, Müller, \& Schäfer, 2021).
For example, Beyens, Pouwels, van Driel, Keijsers, and Valkenburg (2021) reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
This finding is aligned the Differential Susceptibility to Media Effects Model.
Although there is substantial variation of media effects for individual users, the average effects reported in the literature are often small (Valkenburg \& Peter, 2013).

In conclusion, in light of the theoretical considerations and the empirical studies presented above, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.

@@

Achieved via quota sampling, the sample matched the Austrian population in terms of age, gender, region/state, municipality size, and educational level.
In order to participate in the study, the respondents needed to be Austrian residents and had to be at least 14 years of age.
Ethical review and approval was not required for the study in accordance with the local legislation and institutional requirements.
The participants provided their written informed consent to participate in this study.
The average age was 42 years, 49 percent were male, 14 percent had a University degree, and 5 percent were currently unemployed.

@@

In this study, life satisfaction was measured on an 11-point scale.
If people can reliably differentiate 7 levels as mentioned above, this corresponds to 11 / 7 = 1.57 unit change on an 11-point scale.
Hence, a four-point change in media use (e.g., a complete stop) should result in a 1.57-point change in life satisfaction.
In a statistical regression analysis, \emph{b} estimates the change in the dependent variable if the independent variable increases by one point.
We would therefore expect a SESOI of \emph{b} = 1.57 / 4 = 0.39.
For affect, which was measured on a 5-point scale, our SESOI would be \emph{b} = 0.71 / 4 = 0.18.
Because we're agnostic as to whether the effects are positive or negative, the null region includes negative and positive effects.
Finally, in order not to exaggerate precision and to be less conservative, these numbers are reduced to nearby thresholds.\footnote{Note that other researchers also decreased or recommended decreasing thresholds for effect sizes when analyzing within-person or cumulative effects (Beyens et al., 2021; Funder \& Ozer, 2019).}
Together, this leads to a null region ranging from \emph{b} = -.30 to \emph{b} = .30 for life satisfaction, and \emph{b} = -.15 to \emph{b} = .15 for positive and negative affect.

Let's briefly illustrate what this means in practice.
If the 95\% confidence interval falls completely within the null-region (e.g., \emph{b} = .20, {[}95\% CI: .15, .25{]}), the hypothesis that the effect is trivial is supported.
If the confidence interval and the null region overlap (e.g., \emph{b} = .30, {[}95\% CI: .25, .35{]}), the hypothesis is not supported and the results are considered inconclusive, while a meaningful negative effect is rejected.
If the confidence interval falls completely outside of the null-region (e.g., \emph{b} = .40, {[}95\% CI: .35, .45{]}), the hypothesis is rejected and the existence of a meaningful positive effect is supported.
For an illustration, see Figure \ref{fig:sesoi}).

@@

In what follows, I list all the variables that I analyzed.
For the variables' means, range, and variance, see Table \ref{tab:tab-descriptives}.
For a complete list of all items and item characteristics, see \href{https://XMtRA.github.io/Austrian_Corona_Panel_Project}{companion website}.

@@

Life satisfaction was measured with the item ``All things considered, how satisfied are you with your life as a whole nowadays?'' from the European Social Survey (European Social Survey, 2021).
The response options ranged from 0 (\emph{extremely dissatisfied}) to 10 (\emph{extremely satisfied}).
